---
title: "SleepConfidenceIntervals"
output: html_document
---

---
title: "Confidence Intervals with Sleep Data"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Are Students Getting Enough Sleep?

The CDC recommends that adults aged 18-25 get at least 7 hours of sleep per night. A campus health reporter wants to know: are students at this university getting enough sleep?

Remember providing the hours you slept the night before our first class? Your job is to analyze this data and determine what we can confidently say about sleep habits across the entire student body.

This is the core challenge of confidence intervals: using a **sample** to make claims about a **population**.

### Task 1: Load the tidyverse and the data

```{r}
library(tidyverse)

sleep_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/sleep_data.csv")

head(sleep_data)
```

### Task 2: Calculate descriptive statistics

Let's start with what we know about these students. Calculate the mean, standard deviation and number of observations.

```{r}
sleep_stats <- sleep_data |>
  summarise(
    mean_hours = mean(hours),
    sd_hours = sd(hours),
    n = n()
  )

sleep_stats
```

**Answer this question:** Based on these numbers alone, what would you tell the health reporter? Why might it be risky to report the average as the definitive answer for all students?

Although it may appear from this poll that students at this university are, in fact, getting enough sleep, I would caution the health reporter from using the poll to make any kind of definitive statement about student sleep habits because this sample is not nearly large enough to be representative of the whole campus.

### Task 3: Calculate the standard error

The standard error tells us how much our sample mean would vary if we took many different samples of 37 students. It's the key to moving from "this is what 37 students told us" to "this is what we think is true for the whole campus." The formula is: standard deviation divided by the square root of the sample size.

```{r}
sleep_stats <- sleep_stats |>
  mutate(
    se_hours = sd_hours / sqrt(37)
  )

sleep_stats
```

**Answer this question:** The standard error is much smaller than the standard deviation. Why? What does each one measure?

The standard error is measuring how much the sample mean would vary if we took many samples of 37 students. The standard deviation measures the overall spread of the entire dataset. The standard error is smaller because it is only measuring the variation in the average of the dataset.

### Task 4: Calculate a 95% confidence interval

Now we'll build our confidence interval. For a 95% CI with a small sample, we use the t-distribution. The `qt(0.975, df = n - 1)` function gives us the critical value — the number of standard errors we go above and below the mean. We multiply critical value by the standard error to get the margin of error, then add/subtract from the mean.

```{r}
sleep_stats <- sleep_stats |>
  mutate(
    t_critical = qt(0.975, df = n - 1),
    margin_error = t_critical * se_hours,
    ci_lower = mean_hours - se_hours,
    ci_upper = mean_hours + se_hours
  )

sleep_stats |>
  select(mean_hours, margin_error, ci_lower, ci_upper)
```

**Answer this question:** Write one sentence explaining what this confidence interval means in plain English that a reader would understand.

We are 95% confident that the true average of hours of sleep among the total population falls between 7.51 hours and 7.95 hours.

### Task 5: Visualize the distribution with the confidence interval

This chart shows the histogram of individual sleep hours, plus the mean (red line) and the 95% confidence interval (blue shaded region). The shaded region is the range where we think the true average for all students falls.

```{r}
ggplot(sleep_data, aes(x = hours)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", color = "white", alpha = 0.7) +
  geom_vline(xintercept = sleep_stats$mean_hours, color = "red", linewidth = 1) +
  geom_vline(xintercept = sleep_stats$ci_lower, color = "blue", linetype = "dashed") +
  geom_vline(xintercept = sleep_stats$ci_upper, color = "blue", linetype = "dashed") +
  annotate("rect",
           xmin = sleep_stats$ci_lower,
           xmax = sleep_stats$ci_upper,
           ymin = 0, ymax = Inf,
           fill = "blue", alpha = 0.1) +
  annotate("text", x = sleep_stats$mean_hours + 0.1, y = 8,
           label = paste0("Mean: ", round(sleep_stats$mean_hours, 2)),
           hjust = 0, color = "red", fontface = "bold") +
  labs(
    title = "Distribution of Student Sleep Hours",
    subtitle = paste0("95% Confidence Interval for the mean: [",
                     round(sleep_stats$ci_lower, 2), ", ",
                     round(sleep_stats$ci_upper, 2), "]"),
    x = "Hours of Sleep",
    y = "Number of Students"
  ) +
  theme_minimal()
```

**Answer this question:** Look at the chart. Notice that the confidence interval is much narrower than the spread of individual data points. Why? What's the difference between the variation in individual sleep hours and the uncertainty about the average?

The confidence interval is much narrower than the spread of data points because it measures the accuracy of the estimated average hours of sleep, rather than measuring any kind of variability in the individual observations.

### Task 6: The CDC question — is 7 hours within the interval?

The CDC recommends at least 7 hours. Based on the confidence interval, can we say students are getting enough sleep? Let's check.

```{r}
cdc_recommendation <- 7

sleep_stats |>
  mutate(
    meets_cdc = if_else(ci_lower >= cdc_recommendation, "Yes - entire CI is at or above 7", "No - CI includes values below 7")
  ) |>
  select(mean_hours, ci_lower, ci_upper, meets_cdc)
```

**Answer this question:** Based on this result, which of these headlines is justified? Why?

  a) "Students Average More Than 7 Hours of Sleep"
  b) "Students May Not Be Getting Enough Sleep, Survey Suggests"
  c) "Survey: Students Getting Recommended Amount of Sleep"

The third headline is justified because it accurately states that the survey results concluded that students are getting the recommended amount of sleep, while also mentioning the survey, which gives more clarity regarding the extent of the sample. The first headline isn't justified because it doesn't mention the survey, which makes it sound like all students everywhere are averaging more than 7 hours of sleep, rather than just those in the sample or those at this university. The second headline is incorrect because the survey does not show that students aren't getting enough sleep.

### Task 7: What about a bigger sample?

The margin of error depends on the sample size. What if we had surveyed 100 students instead of 37 and got the same mean and standard deviation? Let's recalculate with a hypothetical sample size 100.

```{r}
hypothetical_n <- 100

hypothetical_stats <- tibble(
  mean_hours = sleep_stats$mean_hours,
  sd_hours = sleep_stats$sd_hours,
  n = hypothetical_n,
  se_hours = sd_hours / sqrt(hypothetical_n),
  t_critical = qt(0.975, df = hypothetical_n - 1),
  margin_error = t_critical * se_hours,
  ci_lower = mean_hours - margin_error,
  ci_upper = mean_hours + margin_error
)

# Compare the two intervals side by side
comparison <- tibble(
  sample_size = c(sleep_stats$n, hypothetical_n),
  margin_error = c(sleep_stats$margin_error, hypothetical_stats$margin_error),
  ci_lower = c(sleep_stats$ci_lower, hypothetical_stats$ci_lower),
  ci_upper = c(sleep_stats$ci_upper, hypothetical_stats$ci_upper)
)

comparison
```

**Answer this question:** How did the margin of error change? What does this tell you about the relationship between sample size and certainty?

The margin of error got smaller with the larger sample, which suggests that a larger sample reduces uncertainty.

### Task 8: Journalistic interpretation

You now have all the statistics you need. Based on what you've calculated:

Write a headline and a one-paragraph lede for a campus news story about student sleep habits. Your text should accurately reflect the confidence interval — don't overstate or understate the findings.

HL: UMD students are getting recommended hours of sleep, survey suggests

A recent survey concluded that University of Maryland student respondents get an average of 7.7 hours of sleep, adhering to the CDC recommendation of 7 hours each night.
