---
title: "JOUR405: Statistics for Journalists"
subtitle: "Midterm Exam"
name: "Katherine Schutzman"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# JOUR405: Statistics for Journalists
## Midterm Exam

For this exam, you'll analyze several datasets using R and the statistical concepts we've covered in class. Load the tidyverse before beginning, then complete each task. Write your code in the provided blocks and answer the questions in complete sentences.

## Load the tidyverse

```{r}
library(tidyverse)
```

---

## Part 1: Restaurant Health Inspections (25 points)

The first dataset contains restaurant health inspection scores from Montgomery County. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv`

The dataset includes the name of the establishment, the number of points for critical and non-critical areas, the total points, maximum points possible, the compliance score, and the grade.

### Task 1: Read the data (2 points)

```{r}
restaurants <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/montco_inspections.csv")
```

### Task 2: Calculate the mean, standard deviation, minimum, and maximum of compliance scores. Save the results in a new dataframe called `compliance_stats`. (2 points for each)

```{r}
compliance_stats <- restaurants |> summarize(mean_score=mean(compliance_score), sd_score=sd(compliance_score), min_score=min(compliance_score), max_score=max(compliance_score))
```

### Task 3: Create a histogram of the compliance scores with a vertical line showing the mean. Include a descriptive title and label the x and y axes using `labs()`. (10 points)

```{r}
compliance_stats |> 
  ggplot() + 
  geom_histogram(aes(x = compliance_score), binwidth = 12,)+
  geom_vline(aes(xintercept = mean(compliance_score)), color = "red") + labs(x = "Compliance Score", y = "Number of Restaurants", title = "Restaurant Health Inspection Scores")

```

### Task 4: Write 2–3 sentences interpreting what the standard deviation and histogram tell us about the distribution of compliance scores. What is newsworthy, and what would your next reporting step be? (5 points)

The standard deviation of 5.83, which is relatively low, tells us that the scores are generally clustered around the mean, which we can see is reflected by the histogram. The vast majority of the scores are clustered around the mean at 93.31, showing that most restaurants had high compliance scores. However, the graph is skewed left, with a few restaurants scoring notably lower and dragging the mean down. I would want to look into which restaurants scored significantly lower, especially any that scored outside of two standard deviations below the mean. Is there any kind of pattern in what kind of restaurants get lower scores? Maybe the area that they are in, or the food they serve? I would also want to learn more about the scoring criteria. Is it normal for so many restaurants to get high scores? Are health inspection standards high enough, or are low standards the reason that so many restaurants score well? I think there could be a story either in the shortcomings of the health inspection standards, or in the places that scores concerningly low, compared to the vast majority.

---

## Part 2: Daily Transit Ridership — Z-Scores and Sampling (30 points)

Your editor wants you to identify unusual days in WMATA's daily ridership data that might be worth investigating as story leads. The dataset contains daily bus and rail totals. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv`

### Task 5: Load the WMATA ridership data. (2 points)

```{r}
wmata_daily <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/wmata_daily.csv")
```

### Task 6: Using `summarize()`, calculate the mean and standard deviation for both bus and rail ridership. Save the results in a dataframe called `wmata_stats`. (5 points)

```{r}
wmata_stats <- wmata_daily |> summarize(mean_bus=mean(bus), mean_rail=mean(rail), sd_bus=sd(bus), sd_rail=sd(rail))

wmata_stats
```

### Task 7: Using `mutate()`, add two new columns to `wmata_daily`: `rail_zscore` (the z-score for rail ridership) and `bus_zscore` (the z-score for bus ridership). Save the result as `wmata_zscores`. (8 points)

```{r}
wmata_stats2 <- wmata_daily |> mutate(mean_bus=mean(bus), mean_rail=mean(rail), sd_bus=sd(bus), sd_rail=sd(rail))

wmata_zscores <- wmata_stats2 |> mutate(rail_zscore=(rail - mean_rail)/sd_rail, bus_zscore=(bus - mean_bus)/sd_bus)

```

### Task 8: Filter `wmata_zscores` to show only days where the absolute value of `rail_zscore` is greater than 2 (unusually high or low rail ridership). Arrange by the absolute value of the z-score, with the most extreme first. How many unusual days are there? (8 points)

```{r}
unusual_scores <- wmata_zscores |>
  filter(abs(rail_zscore) > 2) |>
  arrange(abs(rail_zscore))

unusual_scores
```
There are 6 unusual days.

### Task 9: Using `set.seed()`, take a random sample of 91 days from `wmata_daily` (roughly 25% of the data). Calculate the mean and standard deviation for bus and rail ridership in the sample. How do the sample statistics compare to the population statistics you calculated in Task 6? Write 1–2 sentences comparing the values and explaining what this suggests about random sampling. (4 points)

```{r}
set.seed(34)

sample_91 <- wmata_daily |> sample_n(91)


sample_91 |> summarize(mean_bus91=mean(bus), mean_rail91=mean(rail), sd_bus91=sd(bus), sd_rail91=sd(rail))

wmata_stats

```
The means for both bus an rail ridership in the sample and the total population are within a few thousand of each other, which I would say makes the sample an accurate representation of the dataset. The sample mean for bus ridership is higher than that of the true population, and the sample mean for rail ridership is lower. The standard deviations for both the bus and rail ridership are higher for the sample than that of the entire dataset. This is because the standard deviation is set based on the sample's mean, rather than the whole population's mean. Because this sample only contains about 25% of the data, this is an example of the fact that a small sample is less likely to be representative of the total population.

### Task 10: Write 2–3 sentences: What do the unusual days in your z-score analysis suggest as story leads? (3 points)

Five of the six unusual days were on holidays: Thanksgiving, Christmas and New Year's Day in 2024, New Year's Day in 2025. It would make sense for these days to have less travel than usual (their z-scores say the numbers are two standards deviations below the mean) because people often do not travel on holidays, but rather before or after them. Therefore, I don't think those unusual days are particularly newsworthy. Only Jan. 6, 2024, does not fall on a holiday. This day also has an unusually low z-score, so this could have potential as a story lead if there was an event or occurance that day that could explain the low ridership.

---

## Part 3: Maryland Car Theft Rates (25 points)

Your editor has assigned you a story about car thefts in Maryland and wants you to find out which counties have the highest rates. The dataset contains car theft counts for Maryland counties in 2023 along with population figures. Load the data from: `https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv`

### Task 11: Load the car thefts data. (2 points)

```{r}
md_car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/md_car_thefts.csv")
```

### Task 12: Using `mutate()`, add a column called `theft_rate` that calculates the car theft rate per 10,000 residents for each county. Save the result back to `md_car_thefts`. In 1–2 sentences, explain why a per-capita rate is more useful than raw counts for this comparison. (8 points)

```{r}
md_car_thefts <- md_car_thefts |> mutate(theft_rate = ((thefts_2023)/population) *10000)
```

### Task 13: Calculate the median car theft rate and the total number of car thefts statewide. Show the counties that have rates above the median. (7 points)

```{r}
md_car_thefts |> summarize(median(theft_rate))

md_car_thefts |>  filter(theft_rate > 9.580638)


```

### Task 14: Write 2–3 sentences describing what these calculations reveal about the distribution of car thefts in Maryland. What is the lede of a story based on your findings? (8 points)

Because our calculations were based on the number of car thefts per 10,000 residents, I thought rates would be lower in areas with higher populations. While that may be true in some counties, it is not always the case. For example, Prince George's county has the second highest population on this list, but it also has the second highest car theft rate.

LEDE: 2023 car theft rates in Prince George's county far outpace that of other Maryland counties.

---

## Part 4: Data Analysis Scenario (20 points)

A city parks department releases data showing that its 12 community swimming pools average 187 visitors per day during the summer season, with a standard deviation of 42 visitors. Your editor suspects that a handful of pools are dramatically underperforming and wants a story. You also learn that the department's figures are based on a random sample of 60 days per pool rather than a complete count.

Write 3–4 sentences for each of the following questions:

### Question A: Given that the mean is 187 and standard deviation is 42, which statistical measures would you prioritize using and why? (7 points)

A standard deviation of 42 visitors seems like a large amount of variation when the average number of visitors per day is 187. I would prioritize splitting up the data by pool, and recalculating the mean and standard deviation by location. This would allow us to see if certain pools have significantly lower attendance that could be impacting the total mean and standard deviation for all 12 pools. 

### Question B: How would you use z-scores to identify which pools are genuinely underperforming, and why is that more useful than simply ranking pools by raw attendance? (7 points)

Z-scores because that would give more of an indication of how unusual a certain day's attendance is. Z-scores would indicate significantly underperforming and overperforming days in the sample, which we could use to identify which pools fall into either of those categories. This is more useful than simply ranking pools by raw attendance because one pool may generally have less visitors than other, so we can't base a smaller or less attended pool based on the attendance standards of a larger or more well-attended pool. Z-scores are a more standardized way of identifying underperforming pools. 

### Question C: The department collected its attendance figures using a random sample of 60 days per pool rather than tracking every day. What questions would you ask about the sample before relying on these figures for a story? What concerns would you raise in your story? Use statistical terminology where appropriate. (6 points)

I would want to make sure that every summer day had an equal chance of being selected for the sample. It would also be important to check the weather conditions on the days that each pool was sampled on, because if one pool was samples on multiple days when it was hot and sunny out while another pool was sampled on days when it was raining, the data may not be representative of true attendance patterns. I would also want to know exactly what the city parks department considers to be 'summer days' in order to determine what proportion of the summer 60 days is. If 60 days turns out to be a small proportion of the total summer days, I would have less confidence in this sample because small samples are less likely to be representative of true population data.

---

### Submission Instructions
- Save your work frequently
- Make sure all code blocks run without errors
- Provide clear explanations for your choices

Good luck!
