---
title: "HW2 Prince George's County Crime"
---

In this assignment, we will be working with a dataset reporting the number of crimes by month in Prince George's County, Maryland, for every month in 2024. 

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv. You will create a dataframe called `pg_crime` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
pg_crime <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/pg_crime_totals_2024.csv")
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, are any of the categories more important than others? Why or why not? Do any of them make the job of summarizing the data more difficult? Why or why not?

I would argue that crimes like 'assault shooting' or 'homicide' would generally be more newsworthy than other categories that could be less consequential. Accidents, for example, could be as simple as a fender bender, which doesn't warrant much of a story. More impactful crimes, like a shooting or homicide, would likely be of more interest and importance to a larger audience, and therefore more important to cover.
The fact that crimes like 'accidents' are grouped into one category, however, could make it difficult to summarize, because it could include accidents in a wide range of severity.
Much of the data is subsections of different categories of crime. For example, robbery could fall under 'robbery commercial,' 'robbery other,' 'robbery residential' or 'robbery vehicle.' When searching to categorize a certain crime that one is reporting on, these categories are more specific and therefore make it easier to categorize. However, in summarizing the data, it could make it a longer and more complicated process because it may be necessary to spell out each sub-category.

### Task 4: Calculating Averages

Calculate the total number of incidents per month for each category using `mutate()`, calling the new column `total_incidents`. Then calculate the average number of incidents per month for each category of crime, calling the new column `average_incidents`. Assign the result to a new dataframe called `pg_crime_averages`.

```{r}
pg_crime_averages <- pg_crime |> mutate(total_incidents = Jan + Feb + Mar + Apr + May + Jun + Jul + Aug + Sep + Oct + Nov + Dec) |> mutate(average_incidents = (total_incidents / 12))

pg_crime_averages
```

### Task 5: Describe the Averages

Examine the new dataframe, `pg_crime_averages`, by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the average number of incidents per month for each category. Which categories have the highest average number of incidents per month? Which have the lowest? What does this tell you about the pattern of crime in Prince George's County? Are there categories that you have questions about? Why or why not?

On average, accidents were the most common crime throughout the year, with an average of over 481 per month, according to PG crime data. Residential robbery had the lowest average per month. Auto-related crimes are the most frequent on average, according to the data. Accidents, accidents with impound, theft from auto and auto stolen had the four highest averages per month among all the crime categories. I'm not entirely surprised by this because, considering other more complicated crimes on the list, auto-related ones may be easier to pull off and therefore more common. I am concerned and a little confused about the average number of homicides, which comes out to around five per month. That seems like a category of crime that would be more likely to appear in the news, and five per month seems like a high number, so I wonder why I haven't heard more about those events.


### Task 6: Testing a Theory

One theory about crime is that it is more likely to occur in the summer months. Test this theory by creating three new columns using `mutate()` that calculate the percentage change compared to the average for each category of crime for the months of June, July, and August. Call the new columns `june_change`, `july_change`, and `august_change`. Assign the result to a new dataframe called `summer_crime_averages`.

```{r}
summer_crime_averages <- pg_crime_averages |> mutate(june_change = (Jun - average_incidents)/average_incidents * 100 )|> mutate(july_change = (Jul - average_incidents) /average_incidents *100) |> mutate(august_change = (Aug - average_incidents)/average_incidents *100 )

summer_crime_averages
```

### Task 7: Describe the Summer Crime Averages

Look at the `summer_crime_averages` dataframe by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the percentage change in the average number of incidents per month for each category of crime for the months of June, July, and August. Do you see any patterns? What does this tell you about the theory that crime is more likely to occur in the summer months?

Overall, there are more percentage increases than percentage decreases in the average number of crimes per category committed in the summer than over the year as a whole. The average number of homicides in all three summer months, for example, showed a percentage increase from the average throughout the year. The only two crime categories that showed a percentage decrease in the average number committed over the summer versus over the course of the year were B & E other and theft; every other category showed a percenatge increase in the average number of crimes committed in at least one summer month. I would not go so far as to say that almost every crime on this list is more likely to be committed in the summer, because I would have to look at the percentage change from the year-long average to all the other months of the year first. However, there does seem to be a pattern in that many crimes on this list are more likely to occur during the summer, on average.
### Task 8: What's the Story?

Write up a lede and nutgraf of a story focused on the crime data in Prince George's County. It could be from any of the dataframes you've created in this assignment. The lede should be a single sentence that summarizes the most important information in the data. The nutgraf should be a paragraph that provides more context and explains why the data is important.

In Prince George's county, auto-related crimes were the most prominently reported law offenses in 2024, according to recent data. 

Reports of stolen automobiles, theft from automobiles and accidents saw the highest numbers among other crime reports across the county. Including accidents with impound, there were 9,625 reported accidents throughout the year, topping the county crime chart as the most  reported crime in 2024.

### Task 9: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
