---
title: "HW6_Spotify_Songs"
---

In this assignment, you will be working with a dataset of songs played on Spotify in 2020, and you'll be writing code to explore the range of danceability scores in the data, using standard deviation and histograms. You'll also write up your findings.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```

### Task 2: Load the Spotify Songs dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv

```{r}
spotify_songs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/spotify_2020.csv")
```

### Task 3: Calculate the mean and standard deviation of the danceability scores

Write code to calculate the mean and standard deviation of the danceability scores in the Spotify dataset. Display those.

```{r}
spotify_songs |> summarize(mean = mean(danceability), sd = sd(danceability))
```

### Task 4: Create a histogram of the danceability scores

Write code to create a histogram of the danceability scores in the Spotify dataset, based on the example we did in class.

```{r}
spotify_songs |> 
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.25)
```

### Task 5: Add a line to the histogram for the mean danceability score

Take the same code and add a vertical line to the histogram for the mean danceability score, based on the example in class.

```{r}
spotify_songs |> 
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.15) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1)
```


### Task 6: Add two lines to the histogram to show one standard deviation

Take the previous code and add two lines to the histogram for the 1st standard deviation below and above the mean score.

```{r}
spotify_songs |> 
  ggplot() + 
  geom_histogram(aes(x = danceability), binwidth = 0.15) +
  geom_vline(aes(xintercept = mean(danceability)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) - sd(danceability)), color = "blue", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = mean(danceability) + sd(danceability)), color = "blue", linetype = "dashed", size = 1)
```

```{r}
summary(spotify_songs$danceability)
```
### Task 7: Describe our results and pitch a story

Write a few sentences about the distribution of danceability scores in the Spotify dataset. Is it a normal distribution, or is it skewed in one direction? What does the mean and one standard distribution tell us about the songs in the dataset? Draft a short story pitch from this data.

From my observation, it looks like the data is skewed left. The lower bound for outliers is 0.2443, meaning there are multiple outliers on the left side, whereas the upper bound for outliers is 1.1227, meaning there are no outliers on the right side. This reinforces my initial idea that this is not a normal distribution, as the data set is decidedly not symmetrical on either side of the mean. 

The mean tells us that the average danceability of the songs in the data set is 0.6731. The standard deviation, which is 0.1615482, tells us that the danceability score of most songs is within 0.162 of the mean. Because the standard deviation is relatively low, this means the data has low variability.

My story idea for this data set would be to explore how the danceability scores for each song relate to other factors. For example, are danceability and tempo linked at all? Upon glancing at the numbers briefly, I wouldn't say so, but it would be interesting to create graphs for every other factor (tempo, loudness, key, etc.) and organize the graphs by song so we could see if any of that data lines up with danceability. I would consider this to be an interesting study because we may be able determine why some songs are more danceable than others. Or, alternatively, we may be able to find that none of these factors are actually related to danceability. Either way, I think people would be interested in reading about why some songs may be found to be more danceable than others.

### Task 8: Save Your Work and Put it on GitHub

Save this notebook, then commit it to your local repository. Push the changes to your GitHub repository and submit the URL for this file in ELMS.
