---
title: "HW8_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv

```{r}
MMs <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")

```


### Task 3: Sample 5 rows of data from the dataset

```{r}
set.seed(3)

sample_5 <- MMs |> sample_n(5)

```


### Task 4: Calculate the mean age and standard deviation of the sample, using either the blue or white number of M&Ms

```{r}
sample_5 |> summarize(mean_sample=mean(sample_5$red), sd_sample=sd(sample_5$red))

```

### Task 5: Calculate the mean age and standard deviation of the population, using the same measure as in Task 4

```{r}
MMs |> summarize(mean_MMs=mean(MMs$red), sd_MMs=sd(MMs$red))

```


### Task 6: Compare the sample and population statistics

Compare the two sets of means and standard deviations. Are they similar?

The sample mean and population mean are not right next to each other, but they are only about 6 M&Ms apart. The standard deviation is also lower in the sample at 11.7, compared to 16.4 for the population data. The means and standard deviations aren't drastically different, so I think this sample could still be considered a generally accurate representation of the total population, but it also shows the limited potential of a small sample to be entirely representative of a population.

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

Based on the total population data, there were significantly more red M&Ms in each row, on average, which was reflected in the sample data.

Based on both sets of data, it seems that the sample underestimates the standard deviation of the full M&M data set, because the standard deviation is set based on the sample's mean, rather than the whole population's mean. This is also another example of the fact that a small sample is less likely to be representative of the total population. Random sampling does give us the best chance at having a diverse sample, but true representation is still impeded by the small sample size.

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
