---
title: "HW9_CarTheft_CI"
output: html_document
---

---
title: "HW9: Car Theft Confidence Intervals & Margin of Error"
name: REPLACE WITH YOUR NAME
---

Your editor wants to know what a "typical" month of car thefts looks like in Prince George's County. Not just the average — she wants to know how confident you can be in that number. Using monthly figures, confidence intervals and margin of error, you'll come up with an answer and decide whether the data supports a clear story. As with the previous homework, provide answers in this notebook when prompted and you will have to fill in values where you see REPLACE_ME. Get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# LOAD THE TIDYVERSE
library(tidyverse)
```

### Task 2: Load the Prince George's county crime data

```{r}

# Read the data
car_thefts <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/car_thefts_month.csv") |>
  arrange(month)  # Sort by date

# Basic inspection of the data
glimpse(car_thefts)

```

### Task 3: Let's make a chart and describe the results.

We'll make a line chart using geom_line() to try and illustrate any trend over time. What's your initial reaction to the change in car thefts? Is there a trend? If your editor asked you "how many car thefts happen in a typical month?", what would you say based on this chart alone?

Overall, there is a significant drop in car thefts over the course of the chart, which I wouldn't have expected because I don't know if there is any kind of underlying reason for the drop. If my editor were asking for an average number of car thefts in a typical month, I would probably say somewhere around 350-400, but I would clarify that there has been a downward trend over the past three years.

```{r}

car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "darkblue", size = 3) +
  labs(
    title = "Monthly Car Thefts (July 2023 - February 2026)",
    subtitle = "Is there a trend over time?",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold")
  )

```

### Task 4: Calculate the mean, standard error and margin of error

We've got a small number of months, so we need to quantify the uncertainty around our average. We'll calculate the standard error (which measures how much the sample mean could vary), and the margin of error (which gives us the plus/minus range). Replace the REPLACE_ME values based on what we did in class.

```{r}

theft_summary <- car_thefts |>
  summarize(
    mean_thefts = mean(total),
    sd_thefts = sd(total),
    months = n(),
    se_thefts = sd_thefts / sqrt(months),
    margin_error = qt(0.975, df = months - 1) * se_thefts
  )

# Display the results
theft_summary

```

**Answer this question:** Explain in your own words what the margin of error tells us here. If you were writing a story, how would you use this number? For example, how would you complete this sentence: "Prince George's County averages about ___ car thefts per month, plus or minus ___."

The margin of error gives us a reasonable range of uncertainty for a measurement. In this case, Prince George's County averages about 376 car thefts per month, plus or minus 34.

### Task 5: Build the confidence interval

Now let's turn the margin of error into a full confidence interval — the range where we're 95% confident the true average falls. Replace the REPLACE_ME values.

```{r}

theft_summary <- theft_summary |>
  mutate(
    ci_lower = mean_thefts - margin_error,
    ci_upper = mean_thefts + margin_error
  )

theft_summary |>
  select(mean_thefts, margin_error, ci_lower, ci_upper)

```

**Answer this question:** If a county official claimed "car thefts are down to about 300 per month," does your confidence interval support that claim? What about a claim of 350 per month? Explain how the confidence interval helps you evaluate each claim.

Based on this confidence interval, we can be 95% confident that the true average falls between about 341 and 410 car thefts per month. As 300 does not fall within this range, the confidence interval does not support the county official's claim. 350, however, does fall within the range, so that claim would be supported by the confidence interval.

### Task 6: Visualize the data with the confidence interval

Let's make a chart showing the data with the mean and the 95% confidence interval. Does this change anything about your answer from Task 3?

In Task 3, I said I would've guessed the average to be somewhere around 350-400. That range falls within the confidence interval, so my rough estimate was pretty accurate, but the range in which we can be 95% sure the true mean falls is actaully a bit wider than what I had guessed.

```{r}
# Make a chart with the mean and CI
car_thefts |>
ggplot(aes(x = month, y = total)) +
  geom_line(color = "steelblue", alpha = 0.7) +
  geom_point(size = 2) +
  geom_hline(yintercept = theft_summary$mean_thefts, linetype = "dashed", color = "red", size = 1) +
  geom_ribbon(aes(ymin = theft_summary$ci_lower, ymax = theft_summary$ci_upper),
              fill = "red", alpha = 0.2) +
  annotate("text", x = min(car_thefts$month),
           y = theft_summary$mean_thefts + 10,
           label = paste0("Mean: ", round(theft_summary$mean_thefts, 1)),
           hjust = 0, color = "red") +
  labs(
    title = "Monthly Car Thefts with 95% Confidence Interval for the Mean",
    x = "Month",
    y = "Number of Car Thefts"
  ) +
  theme_minimal()

```

### Task 7: The margin of error and sample size

In class we learned that MoE = 1/√n × 100 as a quick approximation for polls. That same principle applies here: more data means a smaller margin of error. Let's see what would happen if we had 40 months of data instead of the number we have, assuming the same mean and standard deviation. Replace the REPLACE_ME value with 40.

```{r}

hypothetical_summary <- tibble(
  mean_thefts = theft_summary$mean_thefts,
  sd_thefts = theft_summary$sd_thefts,
  months = 40,
  se_thefts = sd_thefts / sqrt(months),
  margin_error = qt(0.975, df = months - 1) * se_thefts,
  ci_lower = mean_thefts - margin_error,
  ci_upper = mean_thefts + margin_error
)

# Compare the two
comparison <- tibble(
  scenario = c("Actual data", "40 months"),
  months = c(theft_summary$months, hypothetical_summary$months),
  margin_error = c(theft_summary$margin_error, hypothetical_summary$margin_error),
  ci_lower = c(theft_summary$ci_lower, hypothetical_summary$ci_lower),
  ci_upper = c(theft_summary$ci_upper, hypothetical_summary$ci_upper)
)

comparison

```

**Answer this question:** How did the margin of error and confidence interval change? What does this tell you about the tradeoff between having more data and being able to make precise claims? If you were covering crime trends, how many months of data would you want before writing a definitive story?

As the months increased, the margin of error decreased and the range of our confidence interval decreased. This shows us that the more data we have, and the larger the poll is, the more precise our claim becomes. If I were covering crime trends, I would ideally want at least three or four years of data, which would mean 36-48 months. However, in writing a truly definitive story, I would want data for as many months as possible in order to have a more precise claim.

### Task 8: Reporting with uncertainty

Based on everything you've calculated, answer the following:

1. Write a headline and a one-paragraph lede about car thefts in Prince George's County. Your text should properly reflect the uncertainty in the data — don't overstate or understate what the numbers show.

2. Your editor says "Just tell me if car thefts are going up or down." Using only the confidence interval (no significance testing yet), what's the most accurate answer you can give? What would you need to do next to answer that question more definitively?

HL: Prince George's County sees recent decrease in car thefts, data shows.

From mid-2023 to  present day, the average number of car thefts in Prince George's county has been around 375 per month. Over the past ten recorded months, however, car theft numbers have consistently fallen below that line.


2. I would tell my editor that, in recent months, car thefts are going down because they've been consistently below the mean for nearly a year. To answer that question more definitively, I would need more months of data in order to reevaluate the mean and shorten the range of the confidence interval.


When you are finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
